How not to constructivize cohomology

### 2018-10-03 slides-pssl104.tex
Beginning of PSSL104 slides

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying#2\par}}}

% Workaround for the issue described at
% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}

@@

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=true}

@@

\title{How topos theory can help commutative algebra}
\author{Ingo Blechschmidt}
\date{June 29th, 2018}

\useinnertheme[shadow=true]{rounded}
\useoutertheme[subsection=false]{miniframes}
\setbeamerfont{block title}{size={}}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

\newcommand{\bignumber}[1]{
  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
}

\newenvironment{indentblock}{%
  \list{}{\leftmargin\leftmargin}%
  \item\relax
}{%
  \endlist
}

\newcommand{\hcancel}[5]{%
  \tikz[baseline=(tocancel.base)]{
    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
  }%
}

@@

\setbeamertemplate{frametitle}{%
  \vskip0.7em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
  \end{beamercolorbox}%
}

@@

\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
    % \inserttitle
    \hfill
    \insertframenumber\,/\,\inserttotalframenumber
  \end{beamercolorbox}%
  \vskip0pt%
}

\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}

\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}

@@

%\addtocounter{framenumber}{-1}
%{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{slides-como2018-titlepage-blurred}}
%\begin{frame}[plain]\end{frame}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
\begin{frame}[c]
  \centering
  \medskip
  \hspace*{-3.9em}%

  \hil{How topos theory can help commutative algebra} \\

  \emph{-- an invitation --}

  \bigskip
  \bigskip

  \begin{columns}
    \small
    \begin{column}{0.33\textwidth}
      \centering
      \bignumber{1}
      \smallskip

      \hil{Background on the internal language}
      \medskip
    \end{column}

    \begin{column}{0.33\textwidth}
      \centering
      \bignumber{2}
      \smallskip

      \hil{Applications in commutative algebra}
      \medskip
    \end{column}

    \begin{column}{0.33\textwidth}
      \centering
      \bignumber{3}
      \smallskip

      \hil{The mystery of \mbox{\!\!nongeometric sequents}}
      \medskip
    \end{column}
  \end{columns}

  \scriptsize
  Ingo Blechschmidt (MPI Leipzig) \\
  Toposes in Como \\
  June 29th, 2018
  \par
\end{frame}}

\section{Motivation}

\begin{frame}{Motivating test cases}
  Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
  Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.

  \jnote{1-2}{The two displayed statements are trivial for fields. It is
  therefore natural to try to reduce the general situation to the field
  situation.}

  \jnote{2}{The displayed proofs, which could have been taken from any
  standard textbook on commutative algebra, succeed in this reduction quite
  easily by employing maximal ideals or minimal prime ideals. However, this way
  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
  (for ensuring the existence of a prime ideal and for ensuring that stalks at
  minimal prime ideals are fields) and even the full axiom of choice (for
  ensuring the existence of a minimal prime ideal).

  It therefore doesn't work in the internal universe of most toposes, and in
  any case it obscures explicit computational content: Statements so simple as
  the two displayed ones should admit explicit, computational proofs.

  We'll learn how the internal language of a certain well-chosen topos provides
  a way to perform the reduction in an entirely constructive manner. If so
  desired, the resulting topos-theoretic proofs can be unwound to yield fully
  explicit, topos-free, direct proofs.

  Beautiful constructive proofs can also be found in Richman's note on
  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
  uses of trivial rings} and in the
  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
  Quitté} on constructive commutative algebra.}

  \jnote{3}{We can slightly reduce the requirements of the proof of the first
  statement by employing not a maximal ideal, but a prime ideal. The existence
  of maximal ideals in nontrivial rings is equivalent to the axiom of choice,
  while the existence of prime ideals is equivalent to the weaker Boolean Prime
  Ideal Theorem. However, this improvement doesn't change the main point; the
  proof is still wildly nonconstructive.}

  \jnote{4-}{Grothendieck's generic freeness lemma is an important theorem in
  algebraic geometry, where it is usually stated in the following geometric
  form:
  \begin{indentblock}Let~$X$ be a reduced scheme. Let~$\B$ be
  an~$\O_X$-algebra of finite type. Let~$\M$ be a~$\B$-module of finite type.
  Then over a dense open,
  \begin{enumerate}
  \item[(a)] $\B$ and~$\M$ are locally free as sheaves of~$\O_X$-modules,
  \item[(b)] $\B$ is of finite presentation as a sheaf of~$\O_X$-algebras and
  \item[(c)] $\M$ is of finite presentation as a sheaf of~$\B$-modules.
  \end{enumerate}\end{indentblock}}

  \jnote{5}{All previously known proofs proceed in a series of reduction steps,
  finally culminating in the case where~$A$ is a Noetherian integral domain.
  They are somewhat convoluted (for instance, the proof in the Stacks Project
  is three pages long) and employ several results in commutative algebra which
  have not yet been constructivized.

  Using the internal language of toposes, we will give a short, conceptual and
  constructive proof of Grothendieck's generic freeness lemma. Again, if so
  desired, one can unwind the internal proof to obtain a
  constructive proof which doesn't reference topos theory. The proof obtained
  in this way is still an improvement on the previously known proofs, requiring
  no advanced prerequisites in commutative algebra, and takes
  \fixedhref{https://arxiv.org/abs/1807.01231}{about
  a page}.}

  \only<1-3>{\begin{columns}[t]
  \begin{column}[t]{0.50\textwidth}
    \centering

    \scalebox{0.8}{$\begin{pmatrix}
      \cdot & \cdot \\
      \cdot & \cdot \\
      \cdot & \cdot
      \end{pmatrix}$}
    \vspace*{-1em}

    \begin{varblock}{\textwidth}{A baby application}
      \justifying
      Let~$M$ be a surjective matrix over~$A$ with more rows than columns.
      Then~$1 = 0$ in~$A$.
    \end{varblock}

    \justifying
    \only<2>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{maximal
    ideal} $\mmm$. The matrix is surjective over the field~$A/\mmm$. This is a
    contradiction to basic linear algebra.}

    \only<3>{\textbf{Proof.} \bad{Assume not.} Then there is~a \bad{prime
    ideal} $\ppp$. The matrix is surjective over the
    field~$\mathrm{Quot}(A/\ppp)$. This is a contradiction to basic linear
    algebra.}
  \end{column}

  \begin{column}[t]{0.5\textwidth}
    \centering

    \scalebox{0.8}{$\begin{pmatrix}
      \cdot & \cdot & \cdot & \cdot \\
      \cdot & \cdot & \cdot & \cdot \\
      \cdot & \cdot & \cdot & \cdot
    \end{pmatrix}$}
    \vspace*{-1em}

    \begin{varblock}{\textwidth}{A child application}
      \justifying
      Let~$M$ be an injective matrix over~$A$ with more columns than rows.
      Then~$1 = 0$ in~$A$.
    \end{varblock}

    \justifying
    \visible<2-3>{\textbf{Proof.} \bad{Assume not.} Then there is a \bad{minimal
    prime ideal} $\ppp$. The matrix is injective over the field~$A_\ppp = A[(A
    \setminus \ppp)^{-1}]$. This is a contradiction to basic linear algebra.}
    \end{column}
  \end{columns}}

  \pause
  \pause
  \pause

  \centering

  \includegraphics[height=3em]{generic-freeness}
  \vspace*{-1em}

  \begin{varblock}{0.9\textwidth}{Generic freeness$\phantom{p}$}
    \justifying
    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).

    If~$f = 0$ is the only element of~$A$ such that
    \begin{enumerate}
      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
    \end{enumerate}
    then $1 = 0$ in~$A$.
  \end{varblock}
  \vspace*{-0.5em}

  \pause\center
  \begin{minipage}{0.9\textwidth}
    \textbf{Proof.} See [Stacks Project, Tag 051Q].
  \end{minipage}
\end{frame}

\section{Internal language}

\begin{frame}{The internal language of a topos}
  For any topos~$\E$ and any formula~$\varphi$, we define the meaning of
  \vspace*{-0.5em}
  \[
    \text{``$\E \models \varphi$''} \quad
    \text{(``$\varphi$ holds in the internal universe of~$\E$'')}
  \]

  \vspace*{-1.0em}
  using (Shulman's extension of) the \hil{Kripke--Joyal semantics}.

  \jnote{1}{The internal language of a topos allows to construct objects and
  morphisms of the topos, formulate statements about them and prove such
  statements \emph{in a naive element-based language}. From the internal point
  of view, objects look like sets [more precisely, types]; morphisms look like
  maps; monomorphisms look like injections; epimorphisms
  look like surjections; group objects look like groups; and so on.

  To determine whether a statement~$\varphi$ holds in the internal universe of a
  given topos, one can use the Kripke--Joyal semantics to translate it into an
  ordinary external statement and then check the validity of the external
  translation.

  For instance, in the ef{}fective topos the curious statement~``any function
  $\NN \to \NN$ is computable'' holds, for its external meaning is the
  triviality ``there is a Turing machine which given a Turing machine computing
  some function~$f : \NN \to \NN$ outputs a Turing machine computing~$f$''. In
  contrast, the statement~``any function~$\NN \to \NN$ is either the zero
  function or not'' does not hold in the ef{}fective topos, since its external
  meaning is ``there exists a Turing machine which given a Turing machine
  computing some function~$f : \NN \to \NN$ decides whether~$f$ is the zero
  function or not''.}

  \jnote{2}{Any theorem which has an intuitionistic proof holds in
  the internal universe of any topos. The restriction to intuitionistic logic
  is not due to philosophical concerns; it is a fact of life that only very few
  toposes validate the law of excluded middle (for instance, sheaf toposes over
  discrete topological spaces do if the law of excluded middle is available in
  the metatheory). Luckily, vast amounts of mathematics can be developed in a
  purely intuitionistic setting.

  The internal language machinery itself can be developed in an intuitionistic
  setting.

  The standard internal language of toposes in not enough for our purposes, as
  it misses unbounded quantification (``for all groups'', ``for all rings'')
  and dependent types. Shulman's \fixedhref{https://arxiv.org/abs/1004.3802}{stack
  semantics} offers what we need. No knowledge of stacks is necessary to enjoy
  his paper. Prior work includes
  \fixedhref{https://www.cl.cam.ac.uk/~amp12/papers/polist/polist.pdf}{Polymorphism
  is Set Theoretic, Constructively} by Pitts and
  \fixedhref{http://www.phil.cmu.edu/projects/ast/Papers/Awodey-Butz-Simpson-Streicher-APAL-2013.pdf}{Relating
  first-order set theories, toposes and categories of classes} by Awodey, Butz,
  Simpson and Streicher (obtained independently and published after Shulman).}

  \vspace*{-1em}
  \begin{columns}
    \def\insertblocktitle{}
    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Set \models \varphi$ \\
      ``$\varphi$ holds in the \\ usual sense.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Sh(X) \models \varphi$ \\
      ``$\varphi$ holds continuously.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Eff \models \varphi$ \\
      ``$\varphi$ holds \\ computably.''
      \usebeamertemplate{block end}\end{column}
  \end{columns}
  \medskip

  Any topos supports \hil{mathematical reasoning}:

  \vspace*{-1.5em}
  \begin{hilblock}
    If~$\E \models \varphi$ and if~$\varphi$ entails~$\psi$
    \pointthis{<2>}{intuitionistically}{%
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no axiom of choice},
    then~$\E \models \psi$.
  \end{hilblock}
\end{frame}

\begin{frame}{The Kripke--Joyal semantics of~$\boldsymbol{\Sh(X)}$}
  \small\vspace*{-0.5em}
  Let~$X$ be a topological space. We recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  \mbox{for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
  \footnotesize
  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all open~$V \subseteq U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all open $V \subseteq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  U \models \forall F\_ \varphi(F) &\Ll&
  \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
  U \models \exists F\_ \varphi(F) &\Ll&
  \hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  \varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]

  \jnote{1}{Many interesting sheaves have few global sections, which is why a
  definition such as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point.

  Here is an explicit example of the translation procedure. Let~$\alpha : F
  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader):
  \footnotesize
  \begin{align*}
    & X \models \speak{$\alpha$ is injective} \\[0.2em]
    \Longleftrightarrow\
    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
        $\alpha_U(s) = \alpha_U(t)$ implies $s = t$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}}
\end{frame}

\begin{frame}{Internalizing parameter-dependence}
  \justifying
  Let~$X$ be a space. A continuous family~$(f_x)_{x \in X}$ of continuous
  functions (that is, a continuous function~$f : X \times \RR \to \RR$;
  $f_x(a) = f(x,a)$) induces an endomorphism of the sheaf~$\C$ of continuous
  functions:
  \[
    \bar f : \C \longrightarrow \C,\
    \text{on $U$:}\ s \longmapsto (x \mapsto f_x(s(x))).
  \]

  \vspace*{-1em}

  \begin{itemize}
    \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$. \bigskip
    \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
    \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is
    increasing}$. \bigskip
    \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
    that for each~$i$ there is a} \mbox{continuous function~$s : U_i \to \RR$
    with~$f_x(s(x)) = 0$ for all~$x \in U_i$,} then $\Sh(X) \models \exists s
    \? \RR\_ \bar f(s) = 0$.
  \end{itemize}

  \jnote{1}{This slide, unrelated to commutative algebra or algebraic geometry,
  aims to illustrate one of the basic uses of the internal language of toposes:
  Upgrading any theorem admitting an intuitionistic proof to a
  parameter-dependent version.

  Constructively, there are several non-equivalent forms of the intermediate
  value theorem. The following version doesn't admit an intuitionistic proof:
  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
  \end{indentblock}
  If there was an intuitionistic proof, the statement would hold in any topos,
  so in particular in sheaf toposes over topological spaces. By the
  translations shown on the slide, this
  would amount to the following strengthening of the intermediate value
  theorem: In continuous families of continuous functions, zeros can locally be
  picked continuously. However, this strengthening is invalid, as
  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

  \centering\includegraphics[width=0.1\textwidth]{zeros-in-families-still}}

  \jnote{2}{In contrast, the following version does admit an intuitionistic
  proof. We can therefore interpret it in sheaf toposes over topological
  spaces and thereby obtain the strengthening that in continuous families of
  strictly increasing continuous functions, zeros can locally be picked
  continuously. You are invited to prove this strengthening directly, without
  reference to the internal language.
  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
  reals which is continuous in the usual epsilon/delta sense and which is
  strictly increasing ($a < b$ implies~$g(a) < g(b)$). Assume~$g(-1) < 0 <
  g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) =
  0$.\end{indentblock}}
\end{frame}

\section{The little Zariski topos}

\begin{frame}{The little Zariski topos}
  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
  \begin{enumerate}
    \item the classifying topos of \hil{local localizations} of $A$,
    \item the classifying locale of \hil{prime filters} of $A$,
    \item the locale given by the frame of \hil{radical ideals} of $A$,
    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or
    \item the topos of sheaves over $\Spec(A)$.
  \end{enumerate}
  Its associated topological space of points is the \hil{classical spectrum}
  \[ \{ \fff \subseteq A \,|\, \text{$\fff$ prime filter} \} + \text{Zariski topology}. \]
  It has \hil{enough points} if the Boolean Prime Ideal Theorem holds.

  \footnotesize
  Prime ideal:\, $0 \in \ppp$;\, $x \in \ppp \wedge y \in \ppp \Rightarrow x+y \in \ppp$;\, $1 \not\in \ppp$;\, $xy \in \ppp \Leftrightarrow x \in \ppp \vee y \in \ppp$

  Prime filter:\, $0 \not\in \fff$;\,\,
  $x+y \in \fff \Rightarrow x \in \fff \vee y \in \fff$;
  \hspace*{0.5pt}\,\,
  $1 \in \fff$;\,
  $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$

  \jnote{1}{Any geometric theory has a classifying topos; if the theory under
  consideration is propositional (doesn't have any sorts), then its classifying
  topos can be chosen to be the topos of sheaves over a locale. One can also
  give a direct account of classifying locales, as a pedagogical stepping stone
  to the full theory of classifying toposes.

  The slide contains a small lie: The classical definition of the spectrum of a
  ring is via the set of prime ideals of~$A$, not prime filters. If the law of
  excluded middle is available, there is no difference between these
  definitions since the complement of a prime ideal is a prime filter and vice
  versa.

  One can also consider the classifying locale of prime \emph{ideals} of~$A$.
  Its associated topological space of points is the set of prime ideals
  of~$A$ equipped with the \emph{constructible topology}.

  In an intuitionistic context, any of the (generalized) spaces of items~1--4
  can be adopted as sensible definitions of the spectrum of~$A$. Item~5 is then
  a tautology. The classical definition of the spectrum as a topological space
  doesn't work very well, because verifying the universal property one expects
  of it requires the Boolean Prime Ideal Theorem. Most dramatically, there are
  rings which are not trivial yet have neither prime ideals nor prime filters.
  The classical definition yields in this case the empty space.}
\end{frame}

\begin{frame}{First steps in the little Zariski topos}
  Let~$A$ be a ring. Let~$\fff_0$ be the \hil{generic prime filter} of~$A$; it
  is a subobject of the constant sheaf~$\ull{A}$ of the little Zariski topos.

>>>>

@@

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=true}
\usepackage{multimedia}
\newcommand{\video}[2]{\movie[width=#2,height=#2,autostart,loop,poster]{}{#1}}
\hypersetup{colorlinks=false}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\DeclareSymbolFont{extraup}{U}{zavm}{m}{n}
\DeclareMathSymbol{\varheart}{\mathalpha}{extraup}{86}

@@

\title{How not to constructivize cohomology}
\author{Ingo Blechschmidt}
\date{October 6th, 2018}

\useinnertheme[shadow=true]{rounded}
\setbeamerfont{block title}{size={}}

@@

\newcommand{\A}{\mathcal{A}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

@@

\newcommand{\bignumber}[1]{
  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
}
\newcommand{\bigheart}{\includegraphics{heart}}

\newenvironment{changemargin}[2]{%
  \begin{list}{}{%
    \setlength{\topsep}{0pt}%
    \setlength{\leftmargin}{#1}%
    \setlength{\rightmargin}{#2}%
    \setlength{\listparindent}{\parindent}%
    \setlength{\itemindent}{\parindent}%
    \setlength{\parsep}{\parskip}%
  }%
  \item[]}{\end{list}}

@@

\setbeamertemplate{frametitle}{%
  \vskip1.2em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
  \end{beamercolorbox}%
}

@@

\newcommand{\insertframeextra}{}
\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
    % \inserttitle
    \hfill
    \insertframenumber\insertframeextra\,/\,\inserttotalframenumber
  \end{beamercolorbox}%
  \vskip0pt%
}

\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}
\newcommand{\bad}[1]{\textcolor{red!90}{\textbf{#1}}}

@@

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
\begin{frame}[c]
  \centering

  \bigskip
  \includegraphics[width=0.2\textwidth]{sheaf}
  \bigskip

  \hil{How not to constructivize cohomology}

  \scriptsize
  \textit{-- interruptions welcome at any point --}
  \bigskip

  Ingo Blechschmidt \\
  University of Verona
  \bigskip

  104th Peripatetic Seminar on Sheaves and Logic in Amsterdam \\
  October 6th, 2018
  \par
\end{frame}}

\section{Sheaf cohomology}

\begin{frame}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \centering{\usebeamercolor[fg]{item}{\textbf{\Large Sheaf cohomology}}}\bigskip

      \justifying
      Let~$E$ be a sheaf of modules over a space~$X$.
      Choose an \bad{injective resolution} $0 \to E \to I^0 \to I^1 \to \cdots$.
      Then $H^n(X; E)$ is the $n$-th cohomology of
      \[ (0 \xrightarrow{\phantom{d^0}} I^0(X) \xrightarrow{d^0} I^1(X) \xrightarrow{d^1} \cdots), \]
      that is~$\operatorname{ker}(d^n)\mathop{/}\operatorname{im}(d^{n-1})$.
    %  \begin{align*}
    %    H^n(X; E) &\defeq \text{$n$-th cohomology of $(0 \to I^0(X) \to I^1(X) \to \cdots)$} \\
    %    &\phantom{\vcentcolon}= \operatorname{ker}(I^n(X) \to I^{n+1}(X))\mathop{/}\operatorname{im}(I^{n-1}(X) \to I^n(X))
    %  \end{align*}
    %  is well-defined up to isomorphism.

      \begin{itemize}
        \item The modules~$H^n(X; E)$ are important invariants.
        \item Let~$A$ be a ring. Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X;
        A)$, if~$X$ is locally semiconnected [XXX].
      \end{itemize}
    \end{column}

    \begin{column}{0.6\textwidth}
      \centering{\usebeamercolor[fg]{item}{\textbf{\Large Constructive mathematics}}}\bigskip

      \justifying
    \end{column}
  \end{columns}
\end{frame}
